From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-06 Pi+/paper/finite.tex
Some text in section 3.1

% High-level explanation of technical devices
First, we will define the required notions of free groups and group presentations, and state some of their most
important properties. Then, we introduce our chosen \emph{Coxeter} presentation for $\Sn$. To solve the word problem for
$\Sn$, we will use a rewriting system, with a suitable, well behaved collection of reduction rules corresponding to the
Coxeter presentation. Finally, we describe the normal forms in this rewriting system, using Lehmer codes, and prove the
correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\cref{sec:equivalence}).

@@

Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group} for the full definition).
Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
lots of path constructors corresponding to each group axiom, characterising its path space is difficult.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    % https://q.uiver.app/?q=WzAsNixbMSwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzIsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMSwzLCJcXHRhdV8xIl0sWzIsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
    \adjustbox{scale=\scalef,center}{
      \begin{tikzcd}
        & {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
        {\tau_1\tau_2\tau_1\tau_1\tau_2} && {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
        {\tau_1\tau_2\tau_2} && {\tau_2\tau_2\tau_1} \\
        & {\tau_1}
        \arrow["\braid"', squiggly, from=1-2, to=2-1]
        \arrow["\braid", squiggly, from=1-2, to=2-3]
        \arrow["\cancel"', squiggly, from=2-1, to=3-1]
        \arrow["\cancel"', squiggly, from=3-1, to=4-2]
        \arrow["\cancel", squiggly, from=2-3, to=3-3]
        \arrow["\cancel", squiggly, from=3-3, to=4-2]
      \end{tikzcd}
    }
    \caption{Converging critical pairs}
    \label{fig:critical-pairs-converging}
  \end{subfigure}
  \begin{subfigure}[b]{0.50\textwidth}
    \adjustbox{scale=\scalef,center}{
      % https://q.uiver.app/?q=WzAsMyxbMCwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFsyLDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiYnJhaWQiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFsyLDEsIlxcYnJhaWQqIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XV0=
      \begin{tikzcd}
        {\tau_3\tau_2\tau_3\tau_1} \\
        && {\tau_3\tau_2\tau_1\tau_3} \\
        {\tau_2\tau_3\tau_2\tau_1}
        \arrow["\braid"', squiggly, from=1-1, to=3-1]
        \arrow["\braid", squiggly, from=1-1, to=2-3]
        \arrow["{\longbraid}", squiggly, from=2-3, to=3-1]
      \end{tikzcd}
    }
    \caption{Non-converging critical pairs converging in $\longcox*$}
    \label{fig:critical-pairs-non-converging}
  \end{subfigure}

>>>>

% High-level explanation of technical devices
First, we will define the required notions of free groups and group presentations, and state some of their most
important properties. Then, we introduce our chosen \emph{Coxeter presentation} for $\Sn$. To solve the word problem for
$\Sn$, we will use a rewriting system, with a suitable, well behaved collection of reduction rules corresponding to the
Coxeter presentation. Finally, we describe the normal forms in this rewriting system, using Lehmer codes, and prove the
correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\cref{sec:equivalence}).

@@

Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}).
Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
lots of path constructors corresponding to each group axiom, characterising its path space is difficult.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    % https://q.uiver.app/?q=WzAsNixbMSwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzIsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMSwzLCJcXHRhdV8xIl0sWzIsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
    \adjustbox{scale=\scalef,center}{
      \begin{tikzcd}
        & {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
        {\tau_1\tau_2\tau_1\tau_1\tau_2} && {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
        {\tau_1\tau_2\tau_2} && {\tau_2\tau_2\tau_1} \\
        & {\tau_1}
        \arrow["\braid"', squiggly, from=1-2, to=2-1]
        \arrow["\braid", squiggly, from=1-2, to=2-3]
        \arrow["\cancel"', squiggly, from=2-1, to=3-1]
        \arrow["\cancel"', squiggly, from=3-1, to=4-2]
        \arrow["\cancel", squiggly, from=2-3, to=3-3]
        \arrow["\cancel", squiggly, from=3-3, to=4-2]
      \end{tikzcd}
    }
    \caption{Converging critical pairs}
    \label{fig:critical-pairs-converging}
  \end{subfigure}
  \begin{subfigure}[b]{0.50\textwidth}
    \adjustbox{scale=\scalef,center}{
    % https://q.uiver.app/?q=WzAsMyxbMSwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFszLDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiYnJhaWQiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFsyLDEsIlxcbG9uZ2JyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XV0=
    \begin{tikzcd}
      & {\tau_3\tau_2\tau_3\tau_1} \\
      &&& {\tau_3\tau_2\tau_1\tau_3} \\
      {\tau_2\tau_3\tau_2\tau_1}
      \arrow["braid"', squiggly, from=1-2, to=3-1]
      \arrow["braid", squiggly, from=1-2, to=2-4]
      \arrow["\longbraid", squiggly, from=2-4, to=3-1]
    \end{tikzcd}
    }
    \caption{A non-converging critical pair in $\cox*$ converging in $\longcox*$}
    \label{fig:critical-pairs-non-converging}
  \end{subfigure}

### 2021-07-06 Pi+/paper/reversible1.tex
working on intro
Some text in section 3.1

\section{A Reversible Programming Language}
%% ~3 pages
\label{sec:reversibleone}
\label{sec:reversibletwo}
\label{langeqeq}
\label{sec:informal}

The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic model which we review in this section.

@@

A specialized language for programming with permutations on finite sets can be built using ideas going back to Kelly,
Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
language and the algebraic closure operators become composition operators in the programming language. Once equipped
with a notion of values and types, we get the following syntactic presentation of a programming language for
permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

>>>>

\section{A Reversible Programming Language}
%% ~3 pages
\label{sec:pi}
\label{sec:reversibleone}
\label{sec:reversibletwo}
\label{langeqeq}
\label{sec:informal}

The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic structure which we review in this section.

@@

In a reversible boolean circuit, the number of input bits matches the number of output bits. Circuits can be composed
sequentially, vertically, or horizontally, respectively preserving the number of bits, adding up the number of bits or
multiplying them. So, to program with reversible circuits, all we need to do is make sure that each primitive operation
preserves the number of bits, which is just a natural number. Natural numbers are the free commutative semiring (or,
commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication, so the commutative
semirig identities can be used to design a logic for reversible programming. This inspired the syntax for various
first-order reversible logics and reversible programming
languages~\cite*{sparksSuperstructuralReversibleLogic2014,jamesInformationEffects2012}.

Natural number identities have no computational content. To obtain a computational interpretation of the commutative rig
structure, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence~\cite{jamesInformationEffects2012}. This is a programming language which embodies the computational content
of isomorphisms of finite types, or permutations. On the semantic side, we need to consider the groupoidification of a
commutative rig, that is, a symmetric rig groupoid. A sound semantics for $\PiLang$ in weak rig groupoids was
established in~\cite{caretteComputingSemiringsWeak2016}, and conjectured to be complete.

% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

% A specialised language for programming with permutations on finite sets can be built using ideas going back to Kelly,
% Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
% rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
% Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
% language and the algebraic closure operators become composition operators in the programming language. Once equipped
% with a notion of values and types, we get the following syntactic presentation of a programming language for
% permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
